Human-level control through deep reinforcement learning

### 2018-01-08 article/article.tex
[Article] Finish the article

\usepackage[utf8]{inputenc}
\usepackage[sort, colon]{natbib}
\usepackage{alifexi}
\usepackage{float}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{subcaption}
\usepackage{color}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{commath}

\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{flushend}

@@

  Reinforcement learning is a subarea of machine learning inspired by neuroscience, where learners have to select actions with
  regards to the environment state and receive a reward accordingly. Each agent's objective is to maximize its cumulative reward across its whole lifetime,
  subdivided into episodes. In the traditional framework of reinforcement learning and more specifically in Q-learning,
  researchers usually deal with a small discrete space to represent the values of Q. This is no longer sufficient if one desires to describe the environment state
  as an image of raw pixels. In consequence, we introduce deep Q-learning methods, which have been found to be really effective in mapping raw pixels
  to abstract high-level features. In deep Q-learning, the estimation of Q-values for a given action are produced using deep neural networks,
  which consist of many neural layers stacked on top of each other.
  These techniques have revealed themselves to be able to beat human experts at playing Atari games
  by learning only from visual features. Despite the current progress, learning how to take complex decisions on the basis of high-dimensional visual data remains an
  ongoing challenge. In particular, because of the instability of neural networks, we also explored state-of-the-art dueling architectures to enhance performance
  by separating the high-level representation of states and action values (rewards).

@@

Because of the very high dimensionality of input visual features, it is not reasonable to consider using a table for storing all Q-values.
Instead, one can design a learnable function that maps pixel values to a single row of this table, thus a vector of Q-values whose length is the size
of the action space. For that purpose we used a Deep Q-Network (DQN) to approximate this function, as in the original DQN article~\citep{mnih2015human}, since feedforward Neural Networks
are known to be universal function approximators (Theorem 2 in~\cite{hornik1991approximation}).
A DQN is a particular instance of Convolutional Neural Networks (CNN) and is composed of a stack of neural layers~\citep{lecun1998gradient}. One characteristic of CNNs is the
presence of convolutional filters that maps pixel luminance to more abstract features. Each filter (or kernel) is locally connected to its output, which
allows the convolutional layer to capture the local information of the pixels, contrary to what dense (fully-connected) layers do. This procedure is greatly
inspired by the notion of receptive field introduced by Hubel and Wiesel~\citep{Hubel1962}.
Let $Q(s, a; \theta_i)$ be the Q-value approximation function represented by the neural network, parameterized by its weights $\theta$ at time step $i$.

@@

Finally it has been pointed out by the original DQN conceptors that DQN has an ability to learn by acquiring experience generated from policies
other than its own. \todo{On peut en faire qqch pas vrai ?}

One of the big advantages of deep reinforcement learning is that it relies on a general-purpose learning procedure which allows to make
abstraction of each game's specificities. Although this project focuses on performing on Breakout (a popular Atari game), we made sure
that our agents could be used to learn playing on any Atari game.

@@

We developed our models with the support of the Arcade Learning Environment~\citep{bellemare13arcade}. This gives the opportunity to either
retrieve the environment state as a raw image (for deep reinforcement learning research) or as a state of the RAM (for classic reinforcement learning
applications). Because of the very high dimensionality of raw images, we applied some preprocessing steps on them. These steps are also aimed at
removing some artifacts produced by the Atari architecture. More specifically, the flickering effect implies that some game sprites only appear
in one out of two frames. An easy solution consists in taking the pixelwise maximum value between the current frame and the previous frame.
Then the sample size has been reduced by a factor of three by mapping RGB values to the pixel luminance, using the following formula:
\begin{equation}
    \text{Luminance} = 0.299 \times R + 0.587 \times G + 0.114 \times B.
\end{equation}

@@

Every 4 steps of the Q-learning, the agent builds a minibatch from its memory (either one of the three mechanisms we just mentioned):
this minibatch is used by the DQN to process one step of stochastic gradient descent. In this way the values of Q are estimated iteratively and
refined over time. The deep neural network incorporates minimal prior knowledge about the environment: the only assumption lies in the number of
outputs in the last hidden layer, which has been set as the number of actions. In every Atari game, the action space is discrete, finite and constant over time:
this allows us to determine the output shape of the neural network depending on the game itself. The number of actions made available by the environment
cannot be inferred by the DQN during training phase.

@@

The perceived rewards of the agent have all been reduced to be either $-1$ for a negative reward, $0$ for a session with no reward, and $+1$
for a positive reward, i.e. at any time $t$: $r_t \in \{-1, 0, +1\}$. No intermediate or larger reward is available to the learning algorithm.
This allows to learn without any previous knowledge on the particular game the DQN is trained on. Yet some rewards can be highly different (in Pac-Man likes,
eating a cherry gives 100 points, a banana gives 5000, and eating a ghost gives $2^n \times 100$ with $n$ the number of eaten ghosts in a single run).
This could be improved by adaptive normalization~\citep{van2016learning}.

@@

The interest of the Hubert loss is to bound the gradient of the loss between $-1$ and $+1$, and is claimed to increase stability~\citep{mnih2015human},
yet no further investigation has been performed.

@@

In addition to the original DQN article, we also reimplemented a Dueling Deep Q-Network \citep{DBLP} (DDQN) in which the high-level representation of states
and action values are separated. This shouldn't be confused with the notion of target neural network. Indeed each of our model is copied to a target model
of the same architecture. The DDQN accompanied by its target DDQN, constitute a whole different agent.
For this purpose,the same convolutional architecture as before was kept, but the output of the last convolutionel layer was split in two streams of
equal sizes (each stream input size is the same as the output size of the last convolutional layer), then the resulting tensors have been flattened.
The first stream, which is designed to represent the potential advantage per action, is fully-connected to 6 neurons (one per action), while the second stream
contains a dense representation of the state value and is fully-connected to a single neuron.

@@

Regular experience replay consists in feeding the CNN with uniformly random samples stored in memory. To ensure the agent remembers experience samples $e_t$ that
occured long time ago ($t$ being significantly less than current step), one must allow the replay memory to be really large. Most researchers use a memory size
of 1,000,000 samples, which is too large for most personnal computers (over 16 Gb in the framework of our project). Our second and main contribution consisted in
overcoming this issue by implementing a hierarchical style of memory and showing that it does not jeopardize the performance of Deep Q-networks.

@@

After this phase, the estimates increase and so does the score per episode, until they both reached stability around 11,000,000 steps.

@@

\todo{Parler du clustering.}

@@

@@

One can observe that the DDQN seems to stabilize around an average reward per episode of 1.4 (though more training time would be needed to ensure convergence),
like the DQN with uniform memory. Yet, $Q$ estimates of the DQN end up around $2$ whereas they are only half as big for the DDQN.

\begin{figure*}[!t]
	\vspace{-1.5cm}
	\begin{subfigure}{.47\textwidth}
		\includegraphics[width=\textwidth]{figures/ddqn_uniform_e_scores}
		\subcaption{Total score per episode of the DDQN using uniform memory during its learning phase, smoothed using a moving average of size 1000.}
	\end{subfigure}
	\hfill
	\begin{subfigure}{.47\textwidth}
		\includegraphics[width=\textwidth]{figures/ddqn_uniform_q_values}
		\caption{Estimation of $Q$ by the DDQN using uniform memory, averaged across actions, for each batch during its learning phase and smoothed using a moving average
		with a window of size 5000.}
	\end{subfigure}
	\begin{subfigure}{.47\textwidth}
		\vspace{-.15cm}
		\includegraphics[width=\textwidth]{figures/dqn_prioritized_e_scores}
		\subcaption{Total score per episode of the DQN with prioritized memory during its learning phase, smoothed using a moving average of size 1000.}
	\end{subfigure}
	\hfill
	\begin{subfigure}{.47\textwidth}
		\vspace{-.15cm}
		\includegraphics[width=\textwidth]{figures/dqn_prioritized_q_values}
		\caption{Estimation of $Q$ by the DQN with prioritized memory, averaged across actions, for each batch during its learning phase and smoothed using a moving average
		with a window of size 5000.}
	\end{subfigure}
	\caption{Adaptation of Figure~\ref{fig:dqn uniform} for both DDQN after 16,000,000 training steps (a and b) and DQN with prioritized memory after 7,000,000
	training steps (c and d).\label{fig:DDQN + DQNP}}
\end{figure*}

\todo{Continuer (ici pour ne pas flinguer la mise en page) l'analyse comparative}

@@

\todo{Les CNNs c'est instable, faut du temps et faire du café et bcp de parameter tuning -> solution plus stable ? Modèle linéaire (cf article d'origine) ?}

\todo{Comme pour la prioritized faut mettre à jour les poids de tous les samples à chaque step, c'est compliqué faut optimiser ça}

\todo{LSTM ?}

@@

	\todo{Ajouter ordre de performance des 3 tests}

\section{Acknowledgements}

  \todo{Remercier Lenaerts, Nowe, Google, OpenAI, Markov}

%\newpage
\footnotesize
\bibliographystyle{apalike}
\bibliography{article}

>>>>

\usepackage[utf8]{inputenc}
\usepackage[sort, colon]{natbib}
\usepackage{alifexi}
\usepackage{float}
\usepackage{dblfloatfix}
\usepackage[font=footnotesize,labelfont=bf]{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{color}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{commath}

\usepackage[colorlinks=true,citecolor=black,linkcolor=blue]{hyperref}
\usepackage{flushend}

@@

  Reinforcement learning is a subarea of machine learning inspired by neuroscience where learners have to select actions with
  regards to the environment state and receive a reward accordingly. Each agent's objective is to maximize its cumulative reward across its whole lifetime,
  subdivided into episodes. In the traditional framework of reinforcement learning and more specifically in Q-learning,
  researchers usually deal with a small discrete space to represent the values of Q. This is no longer sufficient if one desires to describe the environment state
  as an image of raw pixels. Hence, we introduce deep Q-learning methods, which have been found to be really effective in mapping raw pixels
  to abstract high-level features. In deep Q-learning, the estimation of Q-values for a given action are produced using deep neural networks,
  which consists in many neural layers stacked on top of each other.
  These techniques have proved to be able to beat human experts at playing Atari games
  by learning only from visual features. Despite the current progress, learning how to take complex decisions on the basis of high-dimensional visual data remains an
  ongoing challenge. In particular, because of the instability of neural networks, we also explored state-of-the-art dueling architectures to enhance performance
  by separating the high-level representation of states and action values (rewards).

@@

Because of the very high dimensionality of input visual features, it is not reasonable to consider using a table for storing all Q-values.
Instead, one can design a learnable function that maps pixel values to a single row of this table, thus a vector of Q-values whose length is the size
of the action space. For that purpose we used a Deep Q-Network (DQN) to approximate this function, as in the original DQN article~\citep{mnih2015human}, since feedforward Neural Networks
are known to be universal function approximators (Theorem 2 in~\cite{hornik1991approximation}).
A DQN is a particular instance of Convolutional Neural Networks (CNN) and is composed of a stack of neural layers~\citep{lecun1998gradient}. One characteristic of CNNs is the
presence of convolutional filters that map pixel luminance to more abstract features. Each filter (or kernel) is locally connected to its output, which
allows the convolutional layer to capture the local information of the pixels, as opposed to what dense (fully-connected) layers do. This procedure is greatly
inspired by the notion of receptive field introduced by Hubel and Wiesel~\citep{Hubel1962}.
Let $Q(s, a; \theta_i)$ be the Q-value approximation function represented by the neural network, parameterized by its weights $\theta$ at time step $i$.

@@

Finally it has been pointed out by the original DQN conceptors that DQN has an ability to learn by acquiring experience generated from policies
other than its own.

One of the big advantages of deep reinforcement learning is that it relies on a general-purpose learning procedure which allows to make
abstraction of each game's specificities. Although this project focuses on performing on Breakout (a popular Atari game),
our agents could be used to learn playing on most Atari game.

@@

We developed our models with the support of the Arcade Learning Environment~\citep{bellemare13arcade}. This gives the opportunity to either
retrieve the environment state as a raw image (for deep reinforcement learning research) or as a state of the RAM (for classic reinforcement learning
applications). Because of the very high dimensionality of raw images, we applied some preprocessing steps on them. These steps are also aimed at
removing some artifacts produced by the Atari architecture. More specifically, the flickering effect implies that some game sprites only appear
in one out of two frames. An easy solution consists in taking the pixelwise maximum value between the current frame and the previous frame.
Then the sample size has been reduced by a factor of three by mapping RGB values to the pixel luminance using the following formula:
\begin{equation}
    \text{Luminance} = 0.299 \times R + 0.587 \times G + 0.114 \times B.
\end{equation}

@@

Every 16 steps of the Q-learning, the agent builds a minibatch from its memory (either one of the three mechanisms we just mentioned):
this minibatch is used by the DQN to process one step of stochastic gradient descent. In this way the values of Q are estimated iteratively and
refined over time. The deep neural network incorporates minimal prior knowledge about the environment: the only assumption lies in the number of
outputs in the last hidden layer, which has been set as the number of actions. In every Atari game, the action space is discrete, finite and constant over time:
this allows us to determine the output shape of the neural network depending on the game itself. The number of actions made available by the environment
cannot be inferred by the DQN during training phase.

@@

The perceived rewards of the agent have all been clipped to be either $-1$ for a negative reward, $0$ for a session with no reward, and $+1$
for a positive reward, i.e. at any time $t$: $r_t \in \{-1, 0, +1\}$. No intermediate or larger reward is available to the learning algorithm.
This allows to learn without any previous knowledge of the particular game the DQN is trained on. Yet some rewards can be highly different (in Pac-Man likes,
eating a cherry gives 100 points, a banana gives 5000, and eating a ghost gives $2^n \times 100$ with $n$ the number of eaten ghosts in a single run).
This could be improved by adaptive normalization~\citep{van2016learning}.

@@

The interest of the Huber loss is to bound the gradient of the loss between $-1$ and $+1$, and is claimed to increase stability~\citep{mnih2015human},
yet no further investigation has been performed.

@@

In addition to the original DQN article, we also reimplemented a Dueling Deep Q-Network \citep{DBLP:journals/corr/WangFL15} (DDQN) in which the high-level
representation of states
and action values are separated. This shouldn't be confused with the notion of target neural network. Indeed each of our model is copied to a target model
of the same architecture. The DDQN accompanied by its target DDQN, constitute a whole different agent.
For this purpose,the same convolutional architecture as before was kept, but the output of the last convolutional layer was split in two streams of
equal sizes (each stream input size is the same as the output size of the last convolutional layer), then the resulting tensors have been flattened.
The first stream, which is designed to represent the potential advantage per action, is fully-connected to 6 neurons (one per action), while the second stream
contains a dense representation of the state value and is fully-connected to a single neuron.

@@

Regular experience replay consists in feeding the CNN with uniformly random samples stored in memory. To ensure the agent remembers experience samples $e_t$ that
occurred long time ago ($t$ being significantly less than current step), one must allow the replay memory to be really large. Most researchers use a memory size
of 1,000,000 samples, which is too large for most personal computers (over 7 Gb in the framework of our project). Our second and main contribution consisted in
overcoming this issue by implementing a hierarchical style of memory and showing that it does not jeopardize the performance of Deep Q-networks.

@@

After this phase, the estimates increase and so does the score per episode, until they both started to settle around 11,000,000 steps. Even though
the score per episode seems to slowly stabilize, the model would need more training time to fully reach stability.

As the $Q$ values computed at each batch are the $Q$ values for the images that the DQN observed at the update step, it is expected to not see
them fully converge: a SGD is performed every 16 steps, independently of the current image. Therefore if the weights update is performed after
a frame where the agent receives positive reward no matter its action (e.g. right before the ball hits a brick) or after a frame where the agent
receives negative reward, no matter his action (e.g. if the ball is about to fall and the paddle is too far away), the $Q$ values will be different.

@@

The states end up clustered by images with similar $Q$ values estimates, so the DQN estimates close $Q$ values for similar images.

@@

It is interesting to notice that the image with the smallest $Q$ estimation is an image with no ball rather than an image with a ball about
to be missed by the agent: the DQN considers that, in terms of reward expectation, having missed a ball is worse than missing a ball. This
reinforces the hypothesis that the DQN did not learn to play FIRE since not restarting the game means no reward.

\begin{figure*}[!t]
	\vspace{-1.5cm}
	\begin{subfigure}{.47\textwidth}
		\includegraphics[width=\textwidth]{figures/ddqn_uniform_e_scores}
		\subcaption{Total score per episode of the DDQN using uniform memory during its learning phase, smoothed using a moving average of size 1000.}
	\end{subfigure}
	\hfill
	\begin{subfigure}{.47\textwidth}
		\includegraphics[width=\textwidth]{figures/ddqn_uniform_q_values}
		\caption{Estimation of $Q$ by the DDQN using uniform memory, averaged across actions, for each batch during its learning phase and smoothed using a moving average
		with a window of size 5000.}
	\end{subfigure}
	\begin{subfigure}{.47\textwidth}
		\includegraphics[width=\textwidth]{figures/dqn_prioritized_e_scores}
		\subcaption{Total score per episode of the DQN with prioritized memory during its learning phase, smoothed using a moving average of size 1000.}
	\end{subfigure}
	\hfill
	\begin{subfigure}{.47\textwidth}
		\includegraphics[width=\textwidth]{figures/dqn_prioritized_q_values}
		\caption{Estimation of $Q$ by the DQN with prioritized memory, averaged across actions, for each batch during its learning phase and smoothed using a moving average
		with a window of size 5000.}
	\end{subfigure}
	\caption{Adaptation of Figure~\ref{fig:dqn uniform} for both DDQN after 16,000,000 training steps (a and b) and DQN with prioritized memory after 7,000,000
	training steps (c and d).\label{fig:DDQN + DQNP}}
\end{figure*}

@@

One can observe that the DDQN seems to stabilize around an average reward per episode of 1.4 (though more training time would be needed to ensure convergence),
like the DQN with uniform memory. Yet, $Q$ estimates of the DQN end up around $2$ whereas they are only half as big for the DDQN, which is the very point of
the DDQN: avoid overoptimistic estimations~\citep{DBLP:journals/corr/WangFL15}.

On the other hand, the addition of prioritized memory annihilates the results by making no score at all: Table~\ref{tab:scores breakout} shows that the performance
is negative, so this agent is worse than random; it avoids the ball. Again, this is partly consistent with the results of~\cite{DBLP:journals/corr/SchaulQAS15}
who also had worse performances on Breakout. Yet, the agent should still be able to learn how to play. This deficiency can also be due to the reduction of
the replay memory size compared to the uniform memory or even to the instability of nonlinear estimators of the $Q$ function~\citep{tsitsiklis1996analysis}

The distribution of scores per game for both DDQN and DQN with prioritized memory are shown in Figure~\ref{fig:scores DDQN + DQNP}.

The distribution of the DDQN is very similar to the distribution of the DQN (if ignoring the score of 219 which must be the occurrence
of a very unlikely event), only slightly shifted towards left, leading to the mode being [20, 30] for the DDQN versus [30, 40] for the DQN.

@@

The results obtained in this study don't pretend to outperform the original results by~\cite{mnih2015human} because the timing to develop and train the models,
and the available computing power were restricted, causing the training to not be extensive and preventing from reproducing the learning with different
parameters or even different initializations.

Furthermore, due to the computational cost of training a single model on different networks, we had to select a single game to test, out of the 49 games that were tested
in the original work. The choice of \textit{Breakout} was finally made because of the performance obtained in the original work (over 10 times the score of the
human player), as well as its simplicity in terms of features. Before making any attempt to improve on the described methods and models, one should first apply
them to a wider variety of games, since the strength of this approach lies in its very general domain of application. Still, the efficiency of this method on
different games, and then different types of gameplay, has been proved in~\cite{mnih2015human}.

When comparing scores obtained by the trained network and human players, it should be noted that even though the input available to both kind of players was the same,
the learning process itself is not comparable. While human players learn in great part by applying knowledge from games they previously played, our agent had to
infer every useful feature from scratch. It could be argued that each game taken individually was not designed to be the first and only game played by a human. In fact
some elementary actions such as resuming a game immediately after a life loss, which would be performed instinctively by any human player, were not always performed by
the model, resulting in many useless frames.

Also, other improvements have been performed to this Deep Reinforcement Learning task since~\citep{mnih2015human}.

Asynchronous advantage actor-critic (A3C) replaces the experience replay algorithm described hereabove with an asynchronous and parallel execution of multiple agents,
each of whom evolves within its own, independent instance of the environment. The experience replay's role of decreasing the bias towards the most recent experience samples
is performed by independently using different exploration policies and thus simultaneously experiencing a wide range of states. Moreover, replacing the experience replay
algorithm additionally allows the use on-policy reinforcement learning methods such as actor-critic). The A3C algorithm also tackles the high hardware requirements of
standard DQN algorithms, achieving better results in less training time (almost linear in the number of parallel agents) using a standard multi-core CPU~\citep{mnih2016asynchronous}

@@

We have implemented a standard DQN as presented in the original paper by Google, and we integrated two different improvements, namely a prioritized memory and a DDQN.
As expected, both the DQN and the DDQN perform at least as well as Google's human tester, but the DQN with prioritized memory failed.

Yet, given the observed results and, the original work, our implementation seems to work, and could be trained on longer period to keep enhancing the results.

\newpage
\footnotesize
\bibliographystyle{apalike}
\bibliography{article}
